---
title: "Combine data frames"
output: 
  html_document:
    df_print: paged
author: "John Favate"
date: "`r Sys.time()`"
---

<style type="text/css">
.main-container {
  max-width: 1200px;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

This document combines results from various programs, as well as form other data sources to a single data frame that will make some downstream processes easier. First, we will prepare each data frame, then combine them at the end.

```{r}
library(tidyverse)
```

### Biocyc information

REL606/7 are annotated at the BioCyc collection of databases. Various files have been downloaded from there and will be added to the data frame. Read in these various files and set column names. These are files that relate K12 and REL names and have other gene descriptions and such. 
```{r}
# read in the various files, chaning some column names in the process
rel_to_k12 <- read_tsv("../../biocyc_files/rel_to_k12_common.txt") 

rel_info <- read_tsv("../../biocyc_files/rel_info.txt")

rel_info_web <- read_tsv("../../biocyc_files/rel_info_web.txt")

k12_info <- read_tsv("../../biocyc_files/k12_info.txt") %>%
  filter(!is.na(`Accession-1`) & !is.na(`Accession-2`))

k12_info_web <- read_tsv("../../biocyc_files/k12_info_web.txt") %>%
  filter(!is.na(`Accession-1`) & !is.na(`Accession-2`))

# change the column names
names(rel_to_k12) <- c("rel_name", "target_id", "k12_name", "k12_accession_1", "k12_accession_2")

names(rel_info) <- c("rel_name", "target_id", "rel_product", "rel_pathways", "rel_reactions")

names(rel_info_web) <- c("rel_url", "target_id")

names(k12_info) <- c("k12_name", "k12_accession_1", "k12_product", "k12_accession_2", "k12_pathways", "k12_reactions")

names(k12_info_web) <- c("k12_url", "k12_accession_1", "k12_accession_2")
```

Various instances of joining to get those into one data frame. This contains the K12 and REL names, other identifiers, and links (which worked at the time this code was written) to the web pages where they come from.
```{r}
full_rel_info <- right_join(rel_info, rel_info_web, by = c("target_id"))

full_k12_info <- right_join(k12_info, k12_info_web, by = c("k12_accession_1", "k12_accession_2"))

b.df <- left_join(rel_to_k12, full_rel_info, by = c("target_id", "rel_name")) %>%
  left_join(., full_k12_info, by = c("k12_name", "k12_accession_1", "k12_accession_2")) %>%
  select(target_id, contains("name"), contains("accession"), contains("product"), contains("pathway"), contains("reaction"), everything())

# save it
write_csv(b.df, "../../data_frames/biocyc_df.csv")

head(b.df)
```

### Kallisto results

The aim is to show for each row (which is a single gene in a single line) the counts and TPMs for replicates, the mean of those, the counts and TPMs in the 4 ancestral replicates, the average of those, for each seqtype. Surprisingly, this is easier done that said.
```{r}
evo.kdf <- read_csv("../../data_frames/table_s1_read_counts.csv") %>%              # read the df in
  filter(grepl("Ara", line)) %>%                                                   # only want evo for this part
  select(-length, -eff_length) %>%                                                 # remove these columns
  pivot_wider(names_from = c(repl, seqtype), values_from = c(est_counts, tpm)) %>% # reshape to a wide format
  mutate(mean_counts_ribo = (est_counts_rep1_ribo + est_counts_rep2_ribo) / 2,     # calculate the various means
         mean_counts_rna = (est_counts_rep1_rna + est_counts_rep2_rna) / 2,
         mean_tpm_ribo = (tpm_rep1_ribo + tpm_rep2_ribo) / 2,
         mean_tpm_rna = (tpm_rep1_rna + tpm_rep2_rna) / 2)

head(evo.kdf)
```

Now for the ancestors. One data frame that contains the individuals
```{r}
anc.reps <- read_csv("../../data_frames/table_s1_read_counts.csv") %>%
  filter(!(grepl("Ara", line))) %>%
  select(-length, -eff_length) %>% 
  pivot_wider(names_from = c(repl, line, seqtype), values_from = c(est_counts, tpm))

head(anc.reps)
```

Then one with the means
```{r}
anc.means <- read_csv("../../data_frames/table_s1_read_counts.csv") %>%
  filter(!(grepl("Ara", line))) %>%
  select(-length, -eff_length) %>% 
  group_by(seqtype, target_id) %>% 
  summarise(anc_mean_tpm = mean(tpm),
            anc_mean_counts = mean(est_counts)) %>% 
  pivot_wider(names_from = seqtype, values_from = c(anc_mean_tpm, anc_mean_counts))

head(anc.means)
```

Join those three together
```{r}
kdf2 <- list(evo.kdf, anc.reps, anc.means) %>% 
  reduce(left_join, by = "target_id") %>% 
  select(line, target_id, contains("counts"), contains("tpm")) # reordering to group TPM and counts

head(kdf2)
```

### DESeq2 results

Again, we want a data frame in wide as opposed to long format, this is easily accomplished with `pivot_wider`. Add "ds_" as a prefix to names beacuse the riborex results have the same column names
```{r}
deseqdf <- read_csv("../../data_frames/table_s2_fold_changes.csv") %>% 
  pivot_wider(names_from = seqtype, values_from = c(where(is.numeric), sig)) %>% 
  rename_if(is.numeric, ~paste0("ds_", .)) %>% 
  rename("ds_sig_ribo" = "sig_ribo", "ds_sig_rna" = "sig_rna")

head(deseqdf)
```

### Riborex results

Do the same thing as with the DESeq results, placing "rr_" in front instead of "ds_"
```{r}
rrdf <- read_csv("../../data_frames/table_s7_riborex_results.csv") %>% 
  rename_if(is.numeric, ~paste0("rr_", .)) %>% 
  rename("rr_sig" = "sig") %>% 
  select(-deleted) # deletions will come the the DESeq df, don't need them here too

head(rrdf)
```


### The large join

We have 4 data frames to be joined, the end result should have `r nrow(kdf2)` rows. 
```{r}
big.df <- list(kdf2, deseqdf, rrdf) %>% 
  reduce(left_join, by = c("line", "target_id")) %>%  # the first 3 are joined by line and target_id
  left_join(b.df, by = "target_id")                   # the last only by target_id

head(big.df)
```

### Addition of more columns

Add a few more columns of information, namely is this a mutator line and is the gene an e coli protein coding gene, an ERCC control, or a tRNA?
```{r}
big.df2 <- big.df %>% 
  mutate(mutator = ifelse(line %in% c("Aar+6", "Ara+3", "Ara-4", "Ara-2", "Ara-1", "Ara-3"), TRUE, FALSE), 
         gene_type = ifelse(grepl("ECB", target_id), "ECB",                                               
                            ifelse(grepl("ERC", target_id), "ERCC", "tRNA")),
         gene_type = ifelse(grepl("ins", rel_name, ignore.case = TRUE) | grepl("ins", k12_name, ignore.case = TRUE), "ins", gene_type))

# I don't like uppercase letters
names(big.df2) <- tolower(names(big.df2))

head(big.df2)
```

Those are all phage proteins.
```{r}
big.df2 %>% 
  filter(gene_type == "ins") %>% 
  pull(rel_product) %>% 
  unique()
```
Save it
```{r}
write_csv(big.df2, "../../data_frames/all_data.csv")
```

```{r}
sessionInfo()
```

