---
title: "GO analysis"
output: 
  html_document:
    df_print: paged
author: "John Favate"
date: "`r Sys.time()`"
---

```{r}
# Prevent printing of warnings and such in the HTML
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(org.EcK12.eg.db)
library(topGO)
```

As in the KEGG analysis, deleted genes get a fold change of -10 and in this case, a p value of 0.01 because that's how genes are picked for this analysis. 
```{r}
all.data <- read_csv("../../data_frames/all_data.csv") %>%
  filter(gene_type == "ECB") %>% 
  mutate(ds_log2foldchange_rna = ifelse(deleted == TRUE, -10, ds_log2foldchange_rna),
         ds_log2foldchange_ribo = ifelse(deleted == TRUE, -10, ds_log2foldchange_ribo),
         ds_padj_rna = ifelse(deleted == TRUE, .01, ds_padj_rna),
         ds_padj_ribo = ifelse(deleted == TRUE, .01, ds_padj_ribo))
```

## topGO 

Get the vectors needed for topGO. The vector itself is whats used to determine interesting genes, 1 if gene is of interest and 0 if it is not. The names of the vector are supposed to be the gene names. 
```{r}
all.down <- all.data %>%
  mutate(sig = ifelse(ds_padj_rna <= .01 &
                        ds_log2foldchange_rna < 0, 1, 0)) %>% # is this a sig down gene
  dplyr::select(line, k12_name, sig, ds_padj_rna) %>% # only need these cols
  filter(!is.na(ds_padj_rna) &
           !is.na(k12_name)) %>% # can't do anything with nameless genes
  split(.$line) %>%
  map(function(x) {
    # sort by increasing p value
    df <- x %>%
      arrange(ds_padj_rna)
    
    # vector of sig or not
    ps <- df$sig
    
    # gene names as vector names
    names(ps) <- df$k12_name
    
    # return a named vector
    return(ps)
  })

# same but for upregulated genes
all.up <- all.data %>%
  mutate(sig = ifelse(ds_padj_rna <= .01 &
                        ds_log2foldchange_rna > 0, 1, 0)) %>%
  dplyr::select(line, k12_name, sig, ds_padj_rna) %>%
  filter(!is.na(ds_padj_rna) & !is.na(k12_name)) %>%
  split(.$line) %>%
  map(function(x) {
    df <- x %>%
      arrange(ds_padj_rna)
    
    ps <- df$sig
    
    names(ps) <- df$k12_name
    
    return(ps)
  })
```

Function to pick significant genes
```{r}
selection <- function(x){x == 1}
```

```{r}
ontologies <- c("BP", "MF", "CC")

names(ontologies) <- ontologies

up.df <- lapply(ontologies, function(x) {
  lapply(all.up, function(y) {
    # generate the go object
    godata <- new(
      "topGOdata",
      ontology = x,
      allGenes = y,
      geneSel = selection,
      annot = annFUN.GO2genes,
      GO2genes = annFUN.org(
        whichOnto = x,
        feasibleGenes = NULL,
        mapping = "org.EcK12.eg.db",
        ID = "symbol"
      ),
      nodeSize = 5
    )
    
    # run fishers test
    results.fisher <-
      runTest(godata, algorithm = "classic", statistic = "fisher")
    
    # run ks test
    results.ks <-
      runTest(godata, algorithm = "classic", statistic = "ks")
    
    # generate a data frame from the results
    go.table <-
      GenTable(godata,
               fisher = results.fisher,
               KS = results.ks,
               topNodes = 65)
  }) %>%
    bind_rows(.id = "line")
}) %>%
  bind_rows(.id = "ontology")
```

Same, but for downregulated genes
```{r}
down.df <- lapply(ontologies, function(x) {
  lapply(all.down, function(y) {
    # generate the go object
    godata <- new(
      "topGOdata",
      ontology = x,
      allGenes = y,
      geneSel = selection,
      annot = annFUN.GO2genes,
      GO2genes = annFUN.org(
        whichOnto = x,
        feasibleGenes = NULL,
        mapping = "org.EcK12.eg.db",
        ID = "symbol"
      ),
      nodeSize = 5
    )
    
    # run fishers test
    results.fisher <-
      runTest(godata, algorithm = "classic", statistic = "fisher")
    
    # run ks test
    results.ks <-
      runTest(godata, algorithm = "classic", statistic = "ks")
    
    # generate a data frame from the results
    go.table <-
      GenTable(godata,
               fisher = results.fisher,
               KS = results.ks,
               topNodes = 65)
  }) %>%
    bind_rows(.id = "line")
}) %>%
  bind_rows(.id = "ontology")
```

```{r}
go.df <- bind_rows("up" = up.df, "down" = down.df, .id = "direction")

names(go.df) <- tolower(names(go.df)) %>%
  str_replace_all(" ", "_")
```

For some reason, there are duplicate category names that have separate go IDs
```{r}
dup.cats <- go.df %>% 
  dplyr::select(ontology, go.id, term) %>% 
  unique() %>% 
  group_by(ontology, term) %>% 
  tally() %>% 
  ungroup() %>% 
  filter(n > 1) %>% 
  arrange(desc(n)) %>% 
  pull(term)
```

I want to know if among categories that are duplicated, is one of the duplicates significant while the other is not? This happens in almost all of them. I'm going to ignore all of those cases and take only those that are NA in the FALSE column, indicating they dont' have a sig/not sig issue.
```{r}
go.df %>%
  filter(term %in% dup.cats) %>%
  dplyr::select(ontology, go.id, term, annotated, fisher) %>%
  split(.$term) %>%
  map(function(x) {
    x %>%
      mutate(sig = fisher <= .05) %>%
      pull(sig) %>%
      table() %>%
      as_tibble()
  }) %>%
  bind_rows(.id = "term") %>%
  pivot_wider(names_from = ".", values_from = n)

# this is as above, but just extracts the values of interest
ok.dubs <- go.df %>%
  filter(term %in% dup.cats) %>%
  dplyr::select(ontology, go.id, term, annotated, fisher) %>%
  split(.$term) %>%
  map(function(x) {
    x %>%
      mutate(sig = fisher <= .05) %>%
      pull(sig) %>%
      table() %>%
      as_tibble()
  }) %>%
  bind_rows(.id = "term") %>%
  pivot_wider(names_from = ".", values_from = n) %>%
  filter(is.na(`FALSE`)) %>%
  pull(term)
```

Find those GO ids, I'm just going to pick the first one in character order
```{r}
remove.these.ids <- go.df %>% 
  filter(term %in% ok.dubs) %>% 
  dplyr::select(go.id, term) %>% 
  arrange(term, go.id) %>% 
  unique() %>% 
  group_by(term) %>% 
  filter(row_number() != 1) %>% 
  ungroup() %>% 
  pull(go.id)
```

Finally, exclude the ones that are duplicated and have varying significance
```{r}
# this is as above, but just extracts the values of interest
bad.dubs <- go.df %>%
  filter(term %in% dup.cats) %>%
  dplyr::select(ontology, go.id, term, annotated, fisher) %>%
  split(.$term) %>%
  map(function(x) {
    x %>%
      mutate(sig = fisher <= .05) %>%
      pull(sig) %>%
      table() %>%
      as_tibble()
  }) %>%
  bind_rows(.id = "term") %>%
  pivot_wider(names_from = ".", values_from = n) %>%
  filter(!is.na(`FALSE`)) %>%
  pull(term)
```

```{r}
go.df2 <- go.df %>% 
  filter(!(term %in% bad.dubs) & !(go.id %in% remove.these.ids)) %>% 
  mutate(sig = fisher <= .05) %>% 
  group_by(direction, ontology, go.id, term) %>% 
  mutate(total_sig = sum(sig)) %>% 
  ungroup() %>% 
  arrange(direction, ontology, desc(total_sig))
```

Save that
```{r}
write_csv(go.df2, "../../data_frames/table_s10_go_results.csv")
```

```{r}
sessionInfo()
```

